Generalized Plackett-Luce Likelihoods

### 2025-01-30 inst/jss4832.tex
instructions as to how to produce the PDF

% -*- mode: noweb; noweb-default-code-mode: R-mode; -*-
\documentclass[article]{jss}

@@

The \code{hyper2} package~\citep{hankin2009,hankin2017} furnishes
computational support for generalized
Plackett-Luce~\citep{plackett1975} likelihood functions.  The
preferred interpretation is a race (as in track and field athletics):
Given six competitors $1-6$, we ascribe to them nonnegative strengths
$p_1,\ldots, p_6$; the probability that $i$ beats $j$ is
$p_i/(p_i+p_j)$.  It is conventional to normalise so that the total
strength is unity, and to identify a competitor with his strength.
Given an order statistic, say $p_1\succ p_2\succ p_3\succ p_4$, the
Plackett-Luce likelihood function would be

@@

\noindent where~$\mathcal{O}$ is a set of observations and~$s$ a
subset of~$\left\{1,2,\ldots,n\right\}$; numbers~$n_s$ are integers
which may be positive or negative.  The \pkg{hyper2} package has the
ability to evaluate such likelihood functions at any point in $S_n$,
thereby admitting a wide range of non-standard nulls such as order
statistics on the $p_i$~\citep{hankin2017}.  It becomes possible to
analyse a wider range of likelihood functions than standard
Plackett-Luce~\citep{turner2020}.  For example, results involving
incomplete order statistics or teams are tractable.  Further, the
introduction of reified entities (monsters) allows one to consider
draws~\citep{hankin2009}, noncompetitive tactics such as
collusion~\citep{hankin2020}, and the phenomenon of team cohesion
wherein the team becomes stronger than the sum of its
parts~\citep{hankin2010}.  Recent versions of the package include
experimental functionality (\code{cheering()}) to investigate the
relaxing of the assumption of conditional independence of the
forward-ranking process.

@@

Existing \code{hyper2} formalism as described by \citet{hankin2017}
cannot represent Equation~\ref{plackettluce}, because
Equation~\ref{hyper2likelihood} uses sets as the indexing elements,
and in this case we need multisets\footnote{Note that the version of
\code{hyper2} presented by \citet{hankin2017} and reviewed
by~\cite{turner2020} used integer-valued sets together with a print
method that used a complicated mapping system from integers to
competitor names.  Current methodology [following commit
  \code{51a8b46}] is to use sets of character strings which represent
the competitors directly; this allows for easier combination of
observations including different competitors.}.  The declarations

@@

The package provides a number of statistical tests on likelihood
functions, modified from~\citet{hankin2017} to work with \code{hyper3}
objects.  For example, we may assess the hypothesis that all three
players are of equal strength [viz $H_0\colon
p_a=p_b=p_c=\frac{1}{3}$]:

@@

A particularly attractive feature of this analysis is that the
Bradley-Terry strengths have direct operational significance: If two
competitors, say Vadlejch and Vesely, were to throw a javelin, then we
would estimate the probability that Vadlejch would throw further than
Vesely at $\displaystyle p_{\mbox{\tiny Vad}}/\left(p_{\mbox{\tiny
Vad}} + p_{\mbox{\tiny Ves}}\right)\simeq 0.74$.  Indeed, from a
training or selection perspective we might follow \cite{hankin2017}
and observe that log-contrasts~\citep{ohagan2004} appear to have
approximately Gaussian likelihood functions for observations of the
type considered here.  Profile log-likelihood curves for log-contrasts
are easily produced by the package, Figure~\ref{profliklogcont}.  We
see that the credible range for~$\log\left(p_{\mbox{\tiny Vad}}/
p_{\mbox{\tiny Ves}}\right)$ includes zero and we have no strong
evidence for these athletes having different (Bradley-Terry)
strengths.

@@

Draws are not considered in the present work but in principle may be
accommodated, either using likelihoods comprising sums of
Plackett-Luce probabilities~\citep{hankin2017}; or the introduction of
a reified draw entity~\citep{hankin2010}.

>>>>

% -*- mode: noweb; noweb-default-code-mode: R-mode; -*-

% Notes:

% To bibtex this file, copy hyper2.bib from the vignettes/ directory
% of the repo (which is under version control).  To create the pdf
% files for the figures, source file "jss4832_replication.R" which
% will create them.  There are slight differences between the JSS
% paper and the manuscript here and there in the citations.

\documentclass[article]{jss}

@@

The \code{hyper2} package~\citep{hankin2017_rmd} furnishes
computational support for generalized
Plackett-Luce~\citep{plackett1975} likelihood functions.  The
preferred interpretation is a race (as in track and field athletics):
Given six competitors $1-6$, we ascribe to them nonnegative strengths
$p_1,\ldots, p_6$; the probability that $i$ beats $j$ is
$p_i/(p_i+p_j)$.  It is conventional to normalise so that the total
strength is unity, and to identify a competitor with his strength.
Given an order statistic, say $p_1\succ p_2\succ p_3\succ p_4$, the
Plackett-Luce likelihood function would be

@@

\noindent where~$\mathcal{O}$ is a set of observations and~$s$ a
subset of~$\left\{1,2,\ldots,n\right\}$; numbers~$n_s$ are integers
which may be positive or negative.  The \pkg{hyper2} package has the
ability to evaluate such likelihood functions at any point in $S_n$,
thereby admitting a wide range of non-standard nulls such as order
statistics on the $p_i$~\citep{hankin2017_rmd}.  It becomes possible to
analyse a wider range of likelihood functions than standard
Plackett-Luce~\citep{turner2020}.  For example, results involving
incomplete order statistics or teams are tractable.  Further, the
introduction of reified entities (monsters) allows one to consider
draws~\citep{hankin2020}, noncompetitive tactics such as
collusion~\citep{hankin2020}, and the phenomenon of team cohesion
wherein the team becomes stronger than the sum of its
parts~\citep{hankin2010}.  Recent versions of the package include
experimental functionality (\code{cheering()}) to investigate the
relaxing of the assumption of conditional independence of the
forward-ranking process.

@@

Existing \code{hyper2} formalism as described by \citet{hankin2017_rmd}
cannot represent Equation~\ref{plackettluce}, because
Equation~\ref{hyper2likelihood} uses sets as the indexing elements,
and in this case we need multisets\footnote{Note that the version of
\code{hyper2} presented by \citet{hankin2017_rmd} and reviewed
by~\cite{turner2020} used integer-valued sets together with a print
method that used a complicated mapping system from integers to
competitor names.  Current methodology [following commit
  \code{51a8b46}] is to use sets of character strings which represent
the competitors directly; this allows for easier combination of
observations including different competitors.}.  The declarations

@@

The package provides a number of statistical tests on likelihood
functions, modified from~\citet{hankin2017_rmd} to work with \code{hyper3}
objects.  For example, we may assess the hypothesis that all three
players are of equal strength [viz $H_0\colon
p_a=p_b=p_c=\frac{1}{3}$]:

@@

A particularly attractive feature of this analysis is that the
Bradley-Terry strengths have direct operational significance: If two
competitors, say Vadlejch and Vesely, were to throw a javelin, then we
would estimate the probability that Vadlejch would throw further than
Vesely at $\displaystyle p_{\mbox{\tiny Vad}}/\left(p_{\mbox{\tiny
Vad}} + p_{\mbox{\tiny Ves}}\right)\simeq 0.74$.  Indeed, from a
training or selection perspective we might follow \cite{hankin2017_rmd}
and observe that log-contrasts~\citep{ohagan2004} appear to have
approximately Gaussian likelihood functions for observations of the
type considered here.  Profile log-likelihood curves for log-contrasts
are easily produced by the package, Figure~\ref{profliklogcont}.  We
see that the credible range for~$\log\left(p_{\mbox{\tiny Vad}}/
p_{\mbox{\tiny Ves}}\right)$ includes zero and we have no strong
evidence for these athletes having different (Bradley-Terry)
strengths.

@@

Draws are not considered in the present work but in principle may be
accommodated, either using likelihoods comprising sums of
Plackett-Luce probabilities~\citep{hankin2017_rmd}; or the introduction of
a reified draw entity~\citep{hankin2010}.
